Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2014-07-20 cordillera-cycle.tex
Add corrections from Arjen and Irina.

  Despite more than a century of geological observations, the Cordilleran ice
  sheet of North America remains poorly understood in terms of its former
  extent, volume and dynamics. Although geomorphological evidence is abundant,
  its complexity is such that whole ice-sheet reconstructions of advance and
  retreat patterns are lacking. Here we use a numerical ice sheet model
  calibrated against field-based evidence to attempt a quantitative
  reconstruction of the Cordilleran ice sheet history through the last glacial
  cycle. A series of simulations is driven by time-dependent temperature
  offsets from six proxy records located around the globe. Although this
  approach reveals large variations in model response to evolving atmospheric
  forcing, all simulations produce two major glaciation events during
  marine isotope stages~4 (61.9--55.4\,ka) and~2
  (29.5--16.9\,ka). The timing of glaciation is
  better reproduced using temperature reconstructions from Greenland and
  Antarctic ice cores than from regional oceanic sediment cores. During most of
  the last glacial cycle, the modelled ice cover is discontinuous and
  restricted to high mountain areas. However, widespread precipitation over the
  Skeena Mountains favours the persistence of a central ice dome throughout the
  glacial cycle. It acts as a nucleation centre before the last glacial maximum
  and hosts the last remains of Cordilleran ice during the
  early Holocene (10.9--9.5\,ka).

@@

During the last glacial cycle, glaciers and ice caps of the North American
Cordillera have been more extensive than today. At the Last Glacial
Maximum (LGM\renote{
    I decided to adopt ``LGM'', which is very common, but not ``CIS'', for the
    sake of readability. This corresponds to abbreviations used in
    \citet{Seguinot.etal.2014} (and now, in the kappa).}),
a continuous blanket of ice, the Cordilleran ice sheet\renote{
    In \citet{Seguinot.etal.2014}, The Cryosphere changed all my occurences of
    Cordilleran (Greenland etc) \textbf{I}ce \textbf{S}heet to lowercase
    \textbf{i}ce \textbf{s}heet. Let's keep it so for now for consistency
    within the thesis (kappa + papers).}
\citep{Dawson.1888}, extended from the Alaska Range in the north to the
North Cascades in the south. In addition, it extended offshore, where it calved
into the Pacific Ocean, and merged with the western margin of its much larger
neighbour, the Laurentide ice sheet, east of the Rocky Mountains
(Fig.~\ref{fig:locmap}).

More than a century of exploration and geological investigations of the
Cordilleran mountains have led to many observations in support of the former
ice sheet
    \citep{Jackson.Clague.1991}.
Despite the lack of documented end moraines offshore, in the zone of confluence
with the Laurentide ice sheet, and in areas swept by the Missoula floods
    \citep{Carrara.etal.1996},
moraines that demarcate the south-western and north-eastern margins provide key
constraints that allow reasonable reconstructions of maximum ice sheet extents
    (\citealp{Prest.etal.1968}; \citealp[Fig. 1.12]{Clague.1989};
     \citealp{Duk-Rodkin.1999};
     \citealp{Booth.etal.2003}; \citealp{Dyke.2004}).
The LGM Cordilleran ice sheet maximum extent was short-lived, as indicated by
field evidence from radiocarbon dating
    \citep{Clague.etal.1980, Clague.1985, Clague.1986, Porter.Swanson.1998,
           Menounos.etal.2008},
cosmogenic exposure dating
    \citep{Stroeven.etal.2010, Stroeven.etal.2014, Margold.etal.2014},
bedrock deformation in response to former ice loads
    \citep{Clague.James.2002, Clague.etal.2005},
and offshore sedimentary records
    \citep{Cosma.etal.2008, Davies.etal.2011}.
However, former ice thicknesses and, therefore, the ice sheet's contribution to
the LGM sea level low stand
    \citep{Carlson.Clark.2012, Clark.Mix.2002}
remain uncertain.

@@

In contrast, evidence for the deglaciation history of the Cordilleran
ice sheet since the LGM is considerable, albeit mostly at a regional scale.
Geomorphological evidence from south-central British Columbia indicates a rapid
deglaciation, including an early emergence of the elevated areas while thin,
stagnant ice still covered the surrounding lowlands
    \citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}.
This model, although credible, may not apply in all areas of the Cordilleran
ice sheet
    \citep{Margold.etal.2013}.
Although solid evidence for late-glacial glacier re-advances have been found in
the Coast, Columbia and Rocky mountains
    \citep{Clague.etal.1997, Friele.Clague.2002, Friele.Clague.2002a,
           Kovanen.2002, Kovanen.Easterbrook.2002, Lakeman.etal.2008,
           Menounos.etal.2008},
it appears to be more sparse than for formerly glaciated regions surrounding
the North Atlantic \aref\footnote{
    Do we need to add a couple of refs. from Britain, Scandinavia, the Alps,
    Eastern America here? Or a review paper? Could yuo help me with this?},
thus allowing for considerable uncertainty concerning
the possibility of a regional late glacial cold reversal.

@@

Although numerical ice sheet modelling has been established as a useful tool to
improve our understanding of the Cordilleran ice sheet
    (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991},
     \citealp{Marshall.etal.2000}),
the ubiquitously mountainous
topography of the region has presented two major challenges to its application.
First, only recent development of numerical ice sheet models and underlying
scientific computing tools \citep{Bueler.Brown.2009, Balay.etal.2014} has
allowed for high-resolution numerical modelling of glaciers and ice sheets on
mountainous terrain
over millenial time scales \citep[e.g.,][]{Golledge.etal.2012}. Second, the
complex
topography of the North American Cordillera also induces strong gradients in
seasonality and distinct patterns of precipition, thus requiring the use of
high-resolution atmospheric forcing fields as an input to the ice sheet model
\citep{Seguinot.etal.2014}.

@@

The simulations presented here were run using the Parallel Ice Sheet Model
(PISM, development version~11b0a7f and stable version~0.6.1), an open source,
finite difference, shallow ice sheet model \citep{PISM-authors.2014}. The model
requires input on basal topography, sea level, geothermal heat flux and
atmospheric\renote{
    Changed here from ``climate'' to ``atmospheric''. Although I have used
    ``climate forcing'' in the previous paper, I now think that ``atmospheric
    forcing'' is more appropriate: ``climate'' may also encompass the oceanic
    component, which is not assessed here.}
forcing. It computes the evolution of ice extent
and thickness over time, the thermal and dynamic\renote{
    Thermal = temperature, dynamic = force + movement. If I meant movement
    alone, I would use ``cinematic''. Here we could use ``ice temperatures,
    stress and deformation (i.e. velocities)'', but that becomes a bit clumsy,
    I feel.}
states of the ice sheet, and the associated lithospheric response.

Basal topography is derived from the ETOPO1 combined topography and bathymetry
dataset with a~resolution of 1\,arc-min \citep{Amante.Eakins.2009}. Sea level
is lowered as a function of time based on the Spectral Mapping Project\renote{
    I had no time to check for later products, but if you have suggestions for
    something more up-to-date, I can try to include it in later simulations
    (everything will have to be re-run anyway).}
\citep[SPECMAP;][]{Imbrie.etal.1989} time scale. Geothermal heat flux
is applied as a constant value of 70\,\unit{mW\,m^{-2}} at 3\,km depth
(Sect.~\ref{sec:icedyn}). Surface mass balance is computed using a positive
degree-day (PDD) model (Sect.~\ref{sec:surface}). Atmospheric forcing is
provided by a monthly climatology from the North American Regional Reanalysis
\citep[NARR;][]{Mesinger.etal.2006} perturbated by time-dependent and
lapse-rate temperature corrections (Sect.~\ref{sec:atm}).

@@

Ice sheet dynamics are typically modelled using a combination of internal
deformation and basal sliding. PISM is a~shallow ice sheet model, which implies
that the balance of stresses is approximated based on their predominant
components. The Shallow Shelf Approximation (SSA) is used as a ``sliding law''
for the Shallow Ice Approximation (SIA) by adding velocity solutions of the
two approximation
\citep[Eqns.~7--9 and 15]{Bueler.Brown.2009, Winkelmann.etal.2011}\renote{
    Arjen, I don't agree with your comment. Reference to equation number(s)
    show the reader where to look. \citet{Winkelmann.etal.2011} is a rather
    long paper that also treats many other aspects of the model. Their SIA+SSA
    approach builds on \citep{Bueler.Brown.2009}, but it is slightly different.
    It is the one in use in PISM, and in my simulations.}.
Ice rheology depends on temperature and water content through an enthalpy
formulation \citep{Aschwanden.etal.2012}. Surface air temperature derived from
the atmospheric forcing (Sect.~\ref{sec:atm}) provides the upper boundary
condition to the ice enthalpy model. Temperature is computed subglacially to
a~depth of 3\,km\renote{
    Yes, I changed this from 1 to 3\,km as we now model longer period of times
    than in the previous paper. The penetration depth of temperature anomalies
    (i.e. skin depth) depends on the period of temperature oscillations at the
    surface. The longer the period (in our case 100\,ka), the deeper the
    temperature change. A good way to estimate what this parameter should be is
    to solve the 1-D temperature diffusion equation in the case of periodic
    surface forcing, using rock heat capacity and rock thermal conductivity
    values similar as in the numerical model. This basically forms the basis
    for my choice of the 3\,km value.},
where it is conditioned by a lower boundary geothermal heat flux of
70\,\unit{mW\,m^{-2}}. Although this uniform value does not account for the
high spatial geothermal variability in the region
\citep{Blackwell.Richards.2004}, it is, on average, representative of available
heat flow measurements. In the low-resolution simulations, the vertical grid
consists of up to 51~enthalpy layers in the ice sheet and 31~temperature layers
in the bedrock. In the high-resolution simulations, up to 101~ice layers, and
61~bedrock layers are used.

A~pseudo-plastic sliding law \citep{Bueler.Pelt.2014} relates the
bed-parallel shear stresses to the sliding velocity\footnote{
    Should I pull in this equation, too?}.
The yield stress, $\tau_c$,
is modelled using the Mohr--Coulomb criterion,
\begin{equation}
   \tau_c = c_0 + N\,\tan{\phi} \,,
\end{equation}
where cohesion, $c_0$, is assumed to be zero. The friction angle, $\phi$,
varies from 15 to 45{\degree}\renote{
    Different from the previous paper, but fairly similar in effect (see
    $\alpha$ and $\delta$ parameter values and corresponding equations in the
    kappa, basal sliding section). These values are is based on a quick
    sensitivity test on Greenland.}.
It is taken to be a piecewise-linear function of modern bed elevation, with
the lowest value below modern sea level (0\,m above sea level, m~a.s.l) and the
highest value above the approximate elevation of highest shorelines
(200\,m~a.s.l), thus accounting for a~weakening of
till associated with the presence of marine sediments. Effective pressure, $N$,
is related to the ice overburden stress, $\rho gh$, and the modelled amount of
subglacial water, using a formula derived from laboratory experiments with till
extracted from an Antarctic ice stream \citep{Tulaczyk.etal.2000,
Bueler.Pelt.2014},
\begin{equation}
    N = \delta \rho gh \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,,
\end{equation}
where $delta$ is chosen as 0.02, $e_0$ is a measured reference void ratio and
$C_c$ a measured compressibility coefficient (Table~\ref{tab:params}). The
amount of water at the base, $W$, varies from zero to $W_{max}=2$\,m, a
threshold above which instantaneous drainage is assumed.
Finally, the bedrock topography responds to ice load\renote{
    I disagree with writing ``the isostatic response''. The main point in
    using this model is that computes a non-isostatic response.}
following a bedrock deformation model that includes point-wise isostasy,
elastic lithosphere flexure and viscous mantle deformation in the lower
half-space\renote{
    I had put an emphasis on semi-infinite as this is, from a mathematical
    point of view, the key assumption of that model, allowing effective
    solution in Fourier domain. This is the whole point of Ed's paper and
    probably the reason why he picked it for PISM. But yes, I guess that
    ``lower half-space'' is fine, too...}
\citep[Table~\ref{tab:params};][]{Lingle.Clark.1985,Bueler.etal.2007}.

@@

The PDD computation accounts for stochastic temperature variations by assuming
a normal temperature distribution of standard deviation $\sigma$ aroung the
expected value $T_m$. It is expressed by an error-function formulation
\citep{Calov.Greve.2005},
\begin{equation}
    \label{eqn:calovgreve}
    \mathrm{PDD} = \int_{t_1}^{t_2} \mathrm{d}t
        \left[\frac{\sigma}{\sqrt{2\pi}}
                \exp\left({-\frac{T_{m}^2}{2\sigma^2}}\right)
              + \frac{T_{m}}{2} \, \mathrm{erfc}
                \left(-\frac{T_{m}}{\sqrt{2}\sigma}\right)\right] \,,
\end{equation}
which is numerically approximated using week-long sub-intervals. In order to
account for the effects of spatial and seasonal variations of temperature
variability \citep{Seguinot.2013}, $\sigma$ is computed from daily temperature
values from the North American Regional Reanalysis
\citep[NARR,][]{Mesinger.etal.2006}, after excluding variability associated
with the seasonal cycle itself \citep[cf.][]{Seguinot.Rogozhina.2014}\renote{
    I may add a figure in the kappa if I find time, but I don't think that we
    should do it here. The subject is too marginal for the present study.}.
Degree-day factors for snow and ice melt are derived from
mass-balance measurements on contemporary glaciers from the Coast Mountains and
Rocky Mountains in British Columbia
\citep[Table~\ref{tab:params};][]{Shea.etal.2009}.

% -----------------------------------------------------------------------------
\subsection{Atmospheric forcing}
\label{sec:atm}
% -----------------------------------------------------------------------------

Atmospheric forcing of the model consists of a present-day monthly climatology%
\renote{
    ``Climatology'' is a common term in meteorology, used to design a set
    of variables aggregated over several decades, and assumed to represent
    some climate state.},
$\{T_{m0}, P_{m0}\}$, where temperatures are modified by offset time series,
${\Delta}T_{TS}$, and lapse-rate corrections, ${\Delta}T_{LR}$:
\begin{align}
    T_m(t, x, y) &= T_{m0}(x, y) + {\Delta}T_{TS}(t)
                    + {\Delta}T_{LR}(t, x, y) \,, \\
    P_m(t, x, y) &= P_{m0}(x, y) \,.
\end{align}

The present-day monthly climatology, $\{T_{m0}, P_{m0}\}$, was computed from
near-surface air temperature and precipitation rate fields from the NARR
\citep{Mesinger.etal.2006}, averaged between 1979 and 2000. Modern climate of the
North American Cordillera is characterised by strong geographic variations in
temperature seasonality, timing of the maximum annual precipitation, and
daily temperature variability (Fig.~\ref{fig:atm}).
Our choice of data from the NARR is motivated by the need for an accurate,
high-resolution precipitation forcing, as identified in a previous sensitivity
study \citep{Seguinot.etal.2014}.

Temperature offset time-series, ${\Delta}T_{TS}$, are derived from
palaeo-temperature proxy records from
the Greenland Ice Core Project \citep[GRIP,][]{Dansgaard.etal.1993}, the
North Greenland Ice Core Project \citep[NGRIP,][]{Andersen.etal.2004},
the European Project for Ice Coring in Antarctica \citep[EPICA,][]
{Jouzel.etal.2007}, the Vostok ice core \citep{Petit.etal.1999}, and Ocean
Drilling Program (ODP) sites 1012 and 1020, both located off the coast of
California \citep{Herbert.etal.2001}. Palaeo-temperatures from the GRIP and
NGRIP
records were calculated from oxygen isotope (\chem{\delta^{18}O}) measurements
using a quadratic equation \citep{Johnsen.etal.1995},
\begin{equation}
    {\Delta}T_{TS}(t) = -11.88 [\chem{\delta^{18}O}(t)
                               -\chem{\delta^{18}O}(0)]
                        -0.1925[\chem{\delta^{18}O}(t)^2
                                -\chem{\delta^{18}O}(0)^2] \,,
\end{equation}
while temperature reconstructions from Antarctic and oceanic cores were
provided as such. All records were scaled linearly (Table~\ref{tab:records}) in
order to simulate realistic and comparable ice extents at the LGM
(Fig.~\ref{fig:timeseries}).

Finally, lapse-rate corrections, ${\Delta}T_{LR}$, are computed as a function
of ice surface elevation, $s$, using the NARR surface geopotential height
invariant field as a reference topography, $b_{ref}$, by
\begin{align}
    {\Delta}T_{LR}(t, x, y) &= -\gamma [s(t, x, y)-b_{ref}] \\
                            &= -\gamma [h(t, x, y)+b(t, x, y)-b_{ref}],
\end{align}

thus accounting for the evolution of ice thickness, ${h=s-b}$\renote{
    Small change in notation to match with the kappa},
on the one hand, and for differences between the basal topography of the ice
flow model, $b$, and the
NARR reference topography, $b_{ref}$, on the other hand. All simulations use an
annual temperature lapse rate of $\gamma = 6\,\unit{{\degree}C\,km^{-1}}$.
In the rest of this paper, we refer to different model runs by the name of the
proxy record used for the palaeo-temperature forcing.

% =============================================================================
\section{Sensitivity to atmospheric forcing time-series}
\label{sec:results}
% =============================================================================

@@

Despite large differences in the input atmospheric forcing
(Fig.~\ref{fig:timeseries}, upper panel), model output presents consistent
features that can be observed across the range of forcing data used. In all
simulations, modelled ice volumes remain relatively low during most of the
glacial cycle, except during two major glacial events which occur between 61.9
and 55.4\,ka during MIS~4, and between 29.5 and 16.9\,ka during MIS~2
(Fig.~\ref{fig:timeseries}, lower panel). An ice volume minimum is
consistently reached between 52.2 and 45.6\,ka during MIS~3. However, the
magnitude and precise timing of these three events depend significantly on the
choice of proxy record used for time-dependent atmospheric forcing
(Table~\ref{tab:extrema})\renote{
    This new table contains modelled ages, volumes and extents at MIS~4--2}.

>>>>

  Despite more than a century of geological observations, the Cordilleran ice
  sheet of North America remains poorly understood in terms of its former
  extent, volume and dynamics. Although geomorphological evidence is abundant,
  its complexity is such that whole ice-sheet reconstructions of advance and
  retreat patterns are lacking. Here we use a numerical ice sheet model
  calibrated against field-based evidence to attempt a quantitative
  reconstruction of the Cordilleran ice sheet history through the last glacial
  cycle. A series of simulations is driven by time-dependent temperature
  offsets from six proxy records located around the globe. Although this
  approach reveals large variations in model response to evolving climate
  forcing, all simulations produce two major glaciation events during
  marine oxygen isotope stages~4 (61.9--55.4\,ka) and~2
  (29.5--16.9\,ka). The timing of glaciation is
  better reproduced using temperature reconstructions from Greenland and
  Antarctic ice cores than from regional oceanic sediment cores. During most of
  the last glacial cycle, the modelled ice cover is discontinuous and
  restricted to high mountain areas. However, widespread precipitation over the
  Skeena Mountains favours the persistence of a central ice dome throughout the
  glacial cycle. It acts as a nucleation centre before the last glacial maximum
  and hosts the last remains of Cordilleran ice during the
  early Holocene (10.9--9.5\,ka).

@@

During the last glacial cycle, glaciers and ice caps of the North American
Cordillera have been more extensive than today. At the Last Glacial
Maximum (LGM\renote{
    I decided to adopt ``LGM'', which is very common, but not ``CIS'', for the
    sake of readability. This corresponds to abbreviations used in
    \citet{Seguinot.etal.2014} (and now, in the kappa).}),
a continuous blanket of ice, the Cordilleran ice sheet\renote{
    In \citet{Seguinot.etal.2014}, The Cryosphere changed all my occurences of
    Cordilleran (Greenland etc) \textbf{I}ce \textbf{S}heet to lowercase
    \textbf{i}ce \textbf{s}heet. Let's keep it so for now for consistency
    within the thesis (kappa + papers).}
\citep{Dawson.1888}, extended from the Alaska Range in the north to the
North Cascades in the south (Fig.~\ref{fig:locmap}).
In addition, it extended offshore, where it calved
into the Pacific Ocean, and merged with the western margin of its much larger
neighbour, the Laurentide ice sheet, east of the Rocky Mountains.

More than a century of exploration and geological investigations of the
Cordilleran mountains have led to many observations in support of the former
ice sheet
    \citep{Jackson.Clague.1991}.
Despite the lack of documented end moraines offshore, in the zone of confluence
with the Laurentide ice sheet, and in areas swept by the Missoula floods
    \citep{Carrara.etal.1996},
moraines that demarcate the northern and south-western margins provide key
constraints that allow reasonable reconstructions of maximum ice sheet extents
    (\citealp{Prest.etal.1968}; \citealp[Fig. 1.12]{Clague.1989};
     \citealp{Duk-Rodkin.1999};
     \citealp{Booth.etal.2003}; \citealp{Dyke.2004}).
The LGM Cordilleran ice sheet maximum extent was short-lived, as indicated by
field evidence from radiocarbon dating
    \citep{Clague.etal.1980, Clague.1985, Clague.1986, Porter.Swanson.1998,
           Menounos.etal.2008},
cosmogenic exposure dating
    \citep{Stroeven.etal.2010, Stroeven.etal.2014, Margold.etal.2014},
bedrock deformation in response to former ice loads
    \citep{Clague.James.2002, Clague.etal.2005},
and offshore sedimentary records
    \citep{Cosma.etal.2008, Davies.etal.2011}.
However, former ice thicknesses and, therefore, the ice sheet's contribution to
the LGM sea level low stand
    \citep{Carlson.Clark.2012, Clark.Mix.2002}
remain uncertain.

@@

In contrast, evidence for the deglaciation history of the Cordilleran
ice sheet since the LGM is considerable, albeit mostly at a regional scale.
Geomorphological evidence from south-central British Columbia indicates a rapid
deglaciation, including an early emergence of the elevated areas while thin,
stagnant ice still covered the surrounding lowlands
    \citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}.
This model, although credible, may not apply in all areas of the Cordilleran
ice sheet
    \citep{Margold.etal.2013}.
Although solid evidence for late-glacial glacier re-advances have been found in
the Coast, Columbia and Rocky mountains
    \citep{Clague.etal.1997, Friele.Clague.2002, Friele.Clague.2002a,
           Kovanen.2002, Kovanen.Easterbrook.2002, Lakeman.etal.2008,
           Menounos.etal.2008},
it appears to be more sparse than for formerly glaciated regions surrounding
the North Atlantic
    \citep[e.g.,][]{Sissons.1979, Lundqvist.1987,
                    Ivy-Ochs.etal.1999, Stea.etal.2011},
thus allowing for considerable uncertainty concerning
the possibility of a regional late glacial cold reversal.

@@

Although numerical ice sheet modelling has been established as a useful tool to
improve our understanding of the Cordilleran ice sheet
    (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991},
     \citealp{Marshall.etal.2000}),
the ubiquitously mountainous
topography of the region has presented two major challenges to its application.
First, only recent development of numerical ice sheet models and underlying
scientific computing tools \citep{Bueler.Brown.2009, Balay.etal.2014} has
allowed for high-resolution numerical modelling of glaciers and ice sheets on
mountainous terrain
over millenial time scales \citep[e.g.,][]{Golledge.etal.2012}. Second, the
complex
topography of the North American Cordillera also induces strong gradients in
seasonality and distinct patterns of precipition, thus requiring the use of
high-resolution climate forcing fields as an input to the ice sheet model
\citep{Seguinot.etal.2014}.

@@

The simulations presented here were run using the Parallel Ice Sheet Model
(PISM, development version~11b0a7f and stable version~0.6.1), an open source,
finite difference, shallow ice sheet model \citep{PISM-authors.2014}. The model
requires input on basal topography, sea level, geothermal heat flux and
climate forcing. It computes the evolution of ice extent
and thickness over time, the thermal and dynamic\renote{
    Thermal = temperature, dynamic = force + movement. If I meant movement
    alone, I would use ``cinematic''. Here we could use ``ice temperatures,
    stress and deformation (i.e. velocities)'', but that becomes a bit clumsy,
    I feel.}
states of the ice sheet, and the associated lithospheric response.

Basal topography is derived from the ETOPO1 combined topography and bathymetry
dataset with a~resolution of 1\,arc-min \citep{Amante.Eakins.2009}. Sea level
is lowered as a function of time based on the Spectral Mapping Project\renote{
    I had no time to check for later products, but if you have suggestions for
    something more up-to-date, I can try to include it in later simulations
    (everything will have to be re-run anyway).}
\citep[SPECMAP;][]{Imbrie.etal.1989} time scale. Geothermal heat flux
is applied as a constant value of 70\,\unit{mW\,m^{-2}} at 3\,km depth
(Sect.~\ref{sec:icedyn}). Surface mass balance is computed using a positive
degree-day (PDD) model (Sect.~\ref{sec:surface}). Climate forcing is
provided by a monthly climatology averaged from 1979 to 2000 from the North
American Regional Reanalysis \citep[NARR;][]{Mesinger.etal.2006},
perturbated by time-dependent and
lapse-rate temperature corrections (Sect.~\ref{sec:atm}).

@@

Ice sheet dynamics are typically modelled using a combination of internal
deformation and basal sliding. PISM is a~shallow ice sheet model, which implies
that the balance of stresses is approximated based on their predominant
components. The Shallow Shelf Approximation (SSA) is used as a ``sliding law''
for the Shallow Ice Approximation (SIA) by adding velocity solutions of the
two approximation
\citep[Eqns.~7--9 and 15]{Bueler.Brown.2009, Winkelmann.etal.2011}\renote{
    Arjen, I don't agree with your comment. Reference to equation number(s)
    show the reader where to look. \citet{Winkelmann.etal.2011} is a rather
    long paper that also treats many other aspects of the model. Their SIA+SSA
    approach builds on \citep{Bueler.Brown.2009}, but it is slightly different.
    It is the one in use in PISM, and in my simulations.}.
Ice rheology depends on temperature and water content through an enthalpy
formulation \citep{Aschwanden.etal.2012}. Surface air temperature derived from
the climate forcing (Sect.~\ref{sec:atm}) provides the upper boundary
condition to the ice enthalpy model. Temperature is computed subglacially to
a~depth of 3\,km\renote{
    Yes, I changed this from 1 to 3\,km as we now model longer period of times
    than in the previous paper. The penetration depth of temperature anomalies
    (i.e. skin depth) depends on the period of temperature oscillations at the
    surface. The longer the period (in our case 100\,ka), the deeper the
    temperature change. A good way to estimate what this parameter should be is
    to solve the 1-D temperature diffusion equation in the case of periodic
    surface forcing, using rock heat capacity and rock thermal conductivity
    values similar as in the numerical model. This basically forms the basis
    for my choice of the 3\,km value.},
where it is conditioned by a lower boundary geothermal heat flux of
70\,\unit{mW\,m^{-2}}. Although this uniform value does not account for the
high spatial geothermal variability in the region
\citep{Blackwell.Richards.2004}, it is, on average, representative of available
heat flow measurements. In the low-resolution simulations, the vertical grid
consists of up to 51~enthalpy layers in the ice sheet and 31~temperature layers
in the bedrock. In the high-resolution simulations, up to 101~ice layers, and
61~bedrock layers are used.

A~pseudo-plastic sliding law,
\begin{equation}
    \label{eqn:pseudoplastic}
    \vec{\tau}_b = -\tau_c \frac{\vec{v}_b}{{v_{th}}^q\,|\vec{v}_b|^{1-q}} \,,
\end{equation}
relates the bed-parallel shear stresses, $\vec{\tau}_b$, to the sliding
velocity, $\vec{v}_b$ (Table~\ref{tab:params}).
The yield stress, $\tau_c$,
is modelled using the Mohr--Coulomb criterion,
\begin{equation}
   \tau_c = c_0 + N\,\tan{\phi} \,,
\end{equation}
where cohesion, $c_0$, is assumed to be zero. The friction angle, $\phi$,
varies from 15 to 45\degree as a piecewise-linear function of modern bed
elevation, with the lowest value occuring below modern sea level (0\,m above
sea level, m~a.s.l) and the highest value occuring above the generalised
elevation of the highest shorelines (200\,m~a.s.l), thus accounting for
a~weakening of till associated with the presence of marine sediments
(cf. \citealp{Martin.etal.2011}, \citealp[supplement]{Aschwanden.etal.2013},
\citealp{PISM-authors.2014}). Effective pressure, $N$,
is related to the ice overburden stress, $\rho gh$, and the modelled amount of
subglacial water, using a formula derived from laboratory experiments with till
extracted from an Antarctic ice stream \citep{Tulaczyk.etal.2000,
Bueler.Pelt.2014},
\begin{equation}
    N = \delta \rho gh \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,,
\end{equation}
where $\delta$ is chosen as 0.02, $e_0$ is a measured reference void ratio and
$C_c$ a measured compressibility coefficient (Table~\ref{tab:params}). The
amount of water at the base, $W$, varies from zero to $W_{max}=2$\,m, a
threshold above which instantaneous drainage is assumed.
Finally, the bedrock topography responds to ice load\renote{
    I disagree with writing ``the isostatic response''. The main point in
    using this model is that computes a non-isostatic response.}
following a bedrock deformation model that includes local isostasy,
elastic lithosphere flexure and viscous mantle deformation in the lower
half-space\renote{
    I had put an emphasis on semi-infinite as this is, from a mathematical
    point of view, the key assumption of that model, allowing effective
    solution in Fourier domain. This is the whole point of Ed's paper and
    probably the reason why he picked it for PISM. But yes, I guess that
    ``lower half-space'' is fine, too...}
\citep[Table~\ref{tab:params};][]{Lingle.Clark.1985,Bueler.etal.2007}.

@@

The PDD computation accounts for stochastic temperature variations by assuming
a normal temperature distribution of standard deviation $\sigma$ aroung the
expected value $T_m$. It is expressed by an error-function formulation
\citep{Calov.Greve.2005},
\begin{equation}
    \label{eqn:calovgreve}
    \mathrm{PDD} = \int_{t_1}^{t_2} \mathrm{d}t
        \left[\frac{\sigma}{\sqrt{2\pi}}
                \exp\left({-\frac{T_{m}^2}{2\sigma^2}}\right)
              + \frac{T_{m}}{2} \, \mathrm{erfc}
                \left(-\frac{T_{m}}{\sqrt{2}\sigma}\right)\right] \,,
\end{equation}
which is numerically approximated using week-long sub-intervals. In order to
account for the effects of spatial and seasonal variations of temperature
variability \citep{Seguinot.2013}, $\sigma$ is computed from daily temperature
values from the NARR, after excluding variability associated
with the seasonal cycle itself \citep[cf.][]{Seguinot.Rogozhina.2014}\renote{
    I may add a figure in the kappa if I find time, but I don't think that we
    should do it here. The subject is too marginal for the present study.}.
Degree-day factors for snow and ice melt are derived from
mass-balance measurements on contemporary glaciers from the Coast Mountains and
Rocky Mountains in British Columbia
\citep[Table~\ref{tab:params};][]{Shea.etal.2009}.

% -----------------------------------------------------------------------------
\subsection{Climate forcing}
\label{sec:atm}
% -----------------------------------------------------------------------------

Climate forcing of the model consists of a present-day monthly climatology%
\renote{
    ``Climatology'' is a common term in meteorology, used to design a set
    of variables aggregated over several decades, and assumed to represent
    some climate state.},
$\{T_{m0}, P_{m0}\}$, where temperatures are modified by offset time series,
${\Delta}T_{TS}$, and lapse-rate corrections, ${\Delta}T_{LR}$:
\begin{align}
    T_m(t, x, y) &= T_{m0}(x, y) + {\Delta}T_{TS}(t)
                    + {\Delta}T_{LR}(t, x, y) \,, \\
    P_m(t, x, y) &= P_{m0}(x, y) \,.
\end{align}

The present-day monthly climatology was computed from
near-surface air temperature and precipitation rate fields from the NARR,
averaged from 1979 to 2000. Modern climate of the
North American Cordillera is characterised by strong geographic variations in
temperature seasonality, timing of the maximum annual precipitation, and
daily temperature variability (Fig.~\ref{fig:atm}).
The use of NARR is motivated by the need for an accurate,
high-resolution precipitation forcing, as identified in a previous sensitivity
study \citep{Seguinot.etal.2014}.

Temperature offset time-series, ${\Delta}T_{TS}$, are derived from
palaeo-temperature proxy records from
the Greenland Ice Core Project \citep[GRIP,][]{Dansgaard.etal.1993}, the
North Greenland Ice Core Project \citep[NGRIP,][]{Andersen.etal.2004},
the European Project for Ice Coring in Antarctica \citep[EPICA,][]
{Jouzel.etal.2007}, the Vostok ice core \citep{Petit.etal.1999}, and Ocean
Drilling Program (ODP) sites 1012 and 1020, both located off the coast of
California \citep{Herbert.etal.2001}. Palaeo-temperatures from the GRIP and
NGRIP
records were calculated from oxygen isotope (\chem{\delta^{18}O}) measurements
using a quadratic equation \citep{Johnsen.etal.1995},
\begin{equation}
    {\Delta}T_{TS}(t) = -11.88 [\chem{\delta^{18}O}(t)
                               -\chem{\delta^{18}O}(0)]
                        -0.1925[\chem{\delta^{18}O}(t)^2
                                -\chem{\delta^{18}O}(0)^2] \,,
\end{equation}
while temperature reconstructions from Antarctic and oceanic cores were
provided as such. All records were scaled linearly (Table~\ref{tab:records}) in
order to simulate comparable ice extents at the LGM  (Table~\ref{tab:extrema})
and realistic outlines (Fig.~\ref{fig:snapshots}).

Finally, lapse-rate corrections, ${\Delta}T_{LR}$, are computed as a function
of ice surface elevation, $s$, using the NARR surface geopotential height
invariant field as a reference topography, $b_{ref}$:
\begin{align}
    {\Delta}T_{LR}(t, x, y) &= -\gamma [s(t, x, y)-b_{ref}] \\
                            &= -\gamma [h(t, x, y)+b(t, x, y)-b_{ref}],
\end{align}

thus accounting for the evolution of ice thickness, ${h=s-b}$\renote{
    Small change in notation to match with the kappa},
on the one hand, and for differences between the basal topography of the ice
flow model, $b$, and the
NARR reference topography, $b_{ref}$, on the other hand. All simulations use an
annual temperature lapse rate of $\gamma = 6\,\unit{K\,km^{-1}}$.
In the rest of this paper, we refer to different model runs by the name of the
proxy record used for the palaeo-temperature forcing.

% =============================================================================
\section{Sensitivity to climate forcing time-series}
\label{sec:results}
% =============================================================================

@@

I AM HERE.

Despite large differences in the input climate forcing
(Fig.~\ref{fig:timeseries}, upper panel), model output presents consistent
features that can be observed across the range of forcing data used. In all
simulations, modelled ice volumes remain relatively low during most of the
glacial cycle, except during two major glacial events which occur between 61.9
and 55.4\,ka during MIS~4, and between 29.5 and 16.9\,ka during MIS~2
(Fig.~\ref{fig:timeseries}, lower panel). An ice volume minimum is
consistently reached between 52.2 and 45.6\,ka during MIS~3. However, the
magnitude and precise timing of these three events depend significantly on the
choice of proxy record used for time-dependent climate forcing
(Table~\ref{tab:extrema})\renote{
    This new table contains modelled ages, volumes and extents at MIS~4--2}.
